Binning Strategies and Related Loss for Large Data

### 2015-02-24 Paper.Rnw
consistency in sections 3.2 and 3.3

The standard and random binning algorithms are easily extendible to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins in $\mathbb{R}^2$. The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as 
%
\begin{eqnarray}\label{standrecbin}
b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
\end{eqnarray}
%
where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively. The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
%
\begin{eqnarray}\label{randrecbin}
b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
\end{eqnarray}
%
where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension. Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.

@@

Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins. Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers. It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function. \ktm{Note this is the quantile function as defined in R; find related literature source.} The bin centers are defined as the midpoint of each consecutive pair of bin boundaries $\{c^\ast_k = (Q_{k-1/K} + Q_{k/K})/2  \text{ for } k=0,\dots,K \}$. The quantile binning function $b^q_C(\cdot) : (c_j) \rightarrow c^\ast_k$ is defined as 
%
\begin{eqnarray}\label{quantbin}
b^q_C(c_j) = (Q_{k-1/K} + Q_{k/K})/2 \text{ where } Q_{k-1/K} < c_j \le Q_{k/K}.
\end{eqnarray}
%
The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in stictly the univariate setting for this manuscript.

>>>>

The standard and random binning algorithms are easily extendible to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_j^\ast)$, with $1 \le j \le N_X \cdot N_Y =J$, where $J$ total number of indexed bins \sout{in $\mathbb{R}^2$} \hh{XXX J is an integer number, definitely not a real-valued vector}. The standard rectangular binning function $b(.) : (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ is defined as 
%
\begin{eqnarray}\label{standrecbin}
b(x_i,y_i) = (b_X(x_i),b_Y(y_i))
\end{eqnarray}
%
where $b_X(x_i)$ and $b_Y(y_i)$ are the univariate standard binning algorithms for the X and Y dimensions respectively. The random rectangular binning function, $b^r(\cdot): (x_i,y_i) \rightarrow (x^\ast_j,y^\ast_j)$ can similarly be defined as
%
\begin{eqnarray}\label{randrecbin}
b^r(x_i,y_i) = (b^r_X(x_i),b^r_Y(y_i))
\end{eqnarray}
%
where $b^r_X(x_i)$ and $b^r_Y(y_i)$ are univariate random binning algorithms for each dimension. Figure~\ref{fig:BinningTutorial} provides an illustration of each binning process extended to a two dimensional situation.

\hh{The relationship between value pair $(x_i, y_i)$ and bin center $(x^\ast_j,y^\ast_j)$ needs to be discussed further - because it is not the same as for the univariate situation. $J$ is defined, as above as a multiple of $N_X$ and $N_Y$, this needs to be properly adjusted in the binning algorithm. It will be some linear relationship of the form $j_X\cdot N_X + j_Y$. }

@@

Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ \hh{shouldn't you go back to the same notation as before? $N_X$, not $J$?  what are the $c_j$ bin centers? before they were $c_j^ast$.} that we wish to divide into quantile based bins. Defining the number of quantile based bins, $K$, results in a set of possible quantile bin centers. It follows that the bin boundaries are $Q_{k/K}$ for $k \in \{1,\dots,K\}$, where $Q_p$ is the the $p^{th}$ empirical quantile of value using the inverse empirical distribution function. \ktm{Note this is the quantile function as defined in R; find related literature source.} \hh{which function exactly? There are multiple.} The bin centers are defined as the midpoint of each consecutive pair of bin boundaries $\{c^\ast_k = (Q_{k-1/K} + Q_{k/K})/2  \text{ for } k=0,\dots,K \}$. The quantile binning function $b^q_C(\cdot) : (c_j) \rightarrow c^\ast_k$ is defined as 
%
\begin{eqnarray}\label{quantbin}
b^q_C(c_j) = (Q_{k-1/K} + Q_{k/K})/2 \text{ where } Q_{k-1/K} < c_j \le Q_{k/K}.
\end{eqnarray}
%
The two dimensional analog for quantile binning is not explored, as this binning strategy will be used in stictly the univariate setting for this manuscript.
